Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-08-26 comments/paper_7bae395_comments_jens.tex
Adding Jason and Jen's comments

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has 
%  - update montages with TAL coordinates
%  - add MAGeT brain diagram, and remove pseudocode
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))

lm_eqn = function(m) {
  l <- list(a = format(coef(m)[1], digits = 2),
      b = format(abs(coef(m)[2]), digits = 2),
      r2 = format(summary(m)$r.squared, digits = 3));
  if (coef(m)[2] >= 0)  {
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
  } else {
    eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l)    
  }

  as.character(as.expression(eq))
}
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable}
\usepackage{titlesec, siunitx, graphicx}
\usepackage[round,authoryear]{natbib}
\usepackage[section]{placeins}  % keep floats in their place.
\usepackage{authblk}            % Listing Author affiliations
\usepackage[hypcap]{caption}

@@

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T }
\newcommand{\FEPdataset}{SZ-FEP }
\newcommand{\gs}{Gold Standard }
\newcommand{\fsl}{FSL }
\newcommand{\freesurfer}{FreeSurfer }
\newcommand{\maper}{MAPER }

% title and authors
\title{Bootstrapping Multi-atlas Hippocampal Segmentation with \mb}
\author[1]{Jon Pipitone}
\author[1]{Matt T. Park}
\author[1]{Julie Winterburn}
\author[1,7]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4]{Jens C. Pruessner}
\author[4]{Martin Lepage}
\author[1,5,7]{Aristotle N. Voineskos}
\author[1,5,6]{M. Mallar Chakravarty}
\author[ ]{Alzheimer's Disease Neuroimaging Initiative}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health, Hospital for Sick Children, Toronto,
ON,Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Brain Imaging Group, Douglas Hospital Research Centre, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[6]{Institute of Biomedical Engineering, University of Toronto, Toronto,
ON, Canada}
\affil[7]{Institute of Medical Science, University of Toronto, Toronto, ON, Canada}

@@

\begin{abstract}
%Neuroimaging research often relies on automated anatomical segmentations of MR
%images of the brain. Current multi-atlas based approaches provide accurate
%segmentations of brain images by propagating manually derived segmentations of
%specific neuroanatomical structures to unlabelled data. These approaches often
%rely on a large number of such manually segmented atlases that take significant
%time and expertise to produce. We present an algorithm for the automatic
%segmentation of the hippocampus that minimizes the number of atlases needed
%while still achieving similar accuracy to multi-atlas approaches.
\end{abstract}

@@

The hippocampus is a neuroanatomical structure situated in the medial temporal
lobe of the brain, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}. In addition to its known functional roles,
the hippocampus is of interest to neuroscientists because it is implicated in
several forms of brain disfunction such as Alzheimer's disease
\citep{Sabuncu2011} and schizophrenia \citep{Narr2004,Karnik-Henry2012}. In
neuroimaging experiments, magnetic resonance images (MRI) are often used for the
identification of the hippocampus. As such, accurate segmentation of the
hippocampus and its subregions in MRI is a necessary first step to better
understand the unique neuroanatomy of subjects. Typically, the gold standard for
neuroanatomical segmentation is manual delineation by an expert human rater.
However with the availability of increasingly large MRI datasets, the time and
expertise required for manual segmentation becomes prohibitive
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}. Further, there is
little agreement between researchers regarding how exactly the hippocampus
should be identified in MRI images \citep{Geuze2004} and this has led to efforts
to create an unified hippocampal segmentation protocol \citep{Jack2011}.

Automated segmentation techniques that are reliable, objective, and reproducible
are a necessary alternative to manual segmentation. In the case of classical
model-based segmentation methods \citep{Haller1997,Csernansky1998}, an MRI atlas
that was previously manually labelled by an expert rater is matched to target
images using nonlinear registration methods. The resulting nonlinear
transformation is applied to the manual labels (i.e. {\em label propagation}) to
move them into the target image space. While this methodology has been used
successfully in several contexts
\citep{Chakravarty2008,Chakravarty2009,Collins1995,Haller1997}, it is limited in
accuracy due to inaccuracies in the estimated nonlinear transformation itself,
partial volume effects in label resampling, and irreconcilable differences
between the neuroanatomy represented within the atlas and target images.

@@

More recently, many groups have been using multiple atlases to improve overall
segmentation accuracy (i.e. multi-atlas segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into the definitive segmentation for the target. In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such as cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations. For example \citet{Collins2010} demonstrated that only
14 atlases, selected based on highest similarity between medial temporal lobe
neuroanatomy as evaluated by normalized mutual information \citep{Studholme1999}
from a library of 80 atlases, were required to achieve accurate segmentations of
the hippocampus. Additionally, several methods have been explored for label
fusion, including the STAPLE algorithm \citep{Warfield2004} that computes a
probabilistic segmentation using an expectation maximization framework from an
set of competing segmentations, or others where a subset of segmentations can be
estimated using metrics such as the sum of squared differences in the regions of
interest to be segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library ranging between 30
\citep{Heckemann2006} and 80 \citep{Collins2010} manually segmented atlases.
This strategy has the main drawback of being inflexible as it does not easily
accommodate varying the definition of the hippocampal anatomy (such as the
commonly used heuristic of subdividing the hippocampus into head, body, and tail
\citep{Poppenk2011,Pruessner2000}). Furthermore, none of these methods have
demonstrated sufficient flexibility to accommodate atlases that are somehow
exceptional such as those derived from serial histological data
\citep{Chakravarty2006,Yelnik2007} or high-resolution MRI data that enables
robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}. Due
to the recent availability of the latter, there has been increased interest in
the use of probabilistic methods for the identification of the hippocampal
subfields on standard T1-weighted images. Our group recently demonstrated that
through use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{MallarChakravarty2012}.
The novelty of this manuscript is the extension of our multi-atlas methodology
to the hippocampus using more than a single input atlas, while simultaneously
limiting the number of possible inputs used during segmentation, and
demonstrating that accurate identification of the hippocampal subfields is
indeed possible using this methodology.

There are few methods that have attempted to perform multi-atlas segmentation
with a limited number of input atlases. The LEAP algorithm is an elegant
modification to the basic multi-atlas strategy \citep{Wolz2010} in which the
atlas library is grown, beginning with a set of manually labelled atlases, and
successively incorporating unlabelled target images after themselves being
labelled using multi-atlas techniques. The sequence in which target images are
labelled is chosen so that the similarity between the atlas images and the
target images is minimised at each step, effectively allowing for deformations
between very dissimilar images to be broken up into sequences of smaller
deformations. Although \citet{Wolz2010} begin with an atlas library of 30 MR
images, this method could theoretically work using a much smaller atlas library.
In their validation, LEAP was used to segment the whole hippocampus in the
ADNI-1 baseline dataset, achieving a mean Dice score of 0.85 with manual
segmentations.

To the best of our knowledge there are two other segmentation methods that
attempt to define the hippocampus and its subfields using standard T1-weighted
data. The first is included with the FreeSurfer package \citep{VanLeemput2009}.
This work is limited as it omits the tail of the hippocampus and the
segmentation protocol has yet to be fully validated. Nonetheless, it
demonstrates that the applicability of hippocampal subfield segmentation using
data from 10 subjects. In the second method, \citet{Yushkevich2009} acquired
and labelled hippocampal subfields on high-resolution MRI data from post-mortem
medial temporal lobe samples. Using nonlinear registration guided by manually
derived hippocampus masks and specific landmarks, they demonstrate accurate
parcellation of hippocampal subfields in unlabelled MRI volumes.

In this paper we undertake a thorough validation of the \mb algorithm for
segmentation of the hippocampus. First, we address the feasibility and accuracy
of whole hippocampus segmentation with a limited number of input atlases by
performing a multi-fold experiment using a subset of the Alzheimer's Disease
Neuroimaging Initiative (ADNI) 1 dataset. Based on the parameters we find in our
experiment, we then validate our algorithm using all of the data available in
the ADNI Complete 1Yr sample and compare our segmentations to the other
segmentations that are available through the ADNI informatics portal. To ensure
that we have not over-fit our parameters to the aging or neurodegenerative
brain, we also apply our segmentations to a dataset of individuals suffering
from first episode psychosis. Finally, we perform a leave-one-out validation
experiment to determine if the subfields can be accurately identified using our
multi-atlas framework.

@@

\subsection{The \mb Algorithm}
In this paper, we will use the term {\em label} to mean any segmentation (manual or 
derived) of an MR image. {\em Label propagation}, or {\em labelling}, is the 
process by which two images are registered and the resulting transformation 
is applied to the labels from one image to bring them into alignment with the 
other image. We will use the term {\em atlas} to mean a manually segmented image,
and the term {\em template} to mean an automatically segmented image (i.e. 
via label propagation). The terms {\em atlas library} and {\em template library} 
describe any set of such images. Additionally, we will use the term {\em target} 
to refer to an unlabelled image that is undergoing segmentation.

The simplest form of multi-atlas segmentation, (which we will call {\em basic 
multi-atlas segmentation}), involves three steps. First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propagated to the target image space. 
Third, the labels are combined into a single label by label fusion
\citep{Heckemann2006, Heckemann2011}. This method is described in detail in
other publications \citep{Collins2010,Heckemann2011,Aljabar2009}.

\mb bootstraps the creation of a large template library given a limited input atlas 
library, and then uses the template library in basic multi-atlas segmentation. 
Images for the template library are selected from a set of input target images, 
either arbitrarily or so as to reflect the neuroanatomy or demographics of the 
target set as a whole (for instance, by sampling equally from cases and controls). 
The template library images are automatically labelled by each of the atlases
via label propagation. Basic multi-atlas segmentation is then conducted using
the template library to segment the entire set of target images (including the
targets whose images are used in the construction of the template library).
Since each template library image has multiple labels (one from each atlas), the
final number of labels to be fused for each target may be quite large (i.e. \#
of atlas $\times$ \# of templates).

@@

The following section describes for experiments conducted to assess the
segmentation quality of the \mb algorithm. The first two experiments assess the
validity of \mb using a cross-validation design. Experiment 1 investigates the
accuracy of \mb whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 tests hippocampal subfield segmentation quality. The last two
experiments assess the validity of the \mb algorithm when applied to different
diseases: first episode schizophrenia (Experiment 3), and Alzheimer's disease
(Experiment 4). Additionally, in experiment 4, we compare \mb segmentations with
those of well-known automated and manual methods and assess segmentation bias.

@@

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a pool
of images and SNT \gs segmentations from the ADNI1 dataset. This form of
cross-validation allows us to rigorously validate a large number of parameter
settings of \mb (atlas and template library sizes, registration algorithm, and
label fusion method) and select the best parameters to use in subsequent
experiments.

\paragraph{\adnidataset dataset} 
69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Semi-automated segmentations of the
left and right whole hippocampi are made available with a subset of ADNI images
\citep{Hsu2002}. These labels have been generated using the SNT tool from
Medtronic Surgical Navigation Technologies, Louisville, CO (see Supplementary
Materials for detailed discussion of the segmentation process used). Throughout
this paper, we will use these segmentations as the \gs for comparison.

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset used was the "\adnidataset" standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel resolution of $1.25 \times 1.25 \times 1.2mm^3$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% ADNI1 X-Validation Dataset Demographics Table
%
<<ADNI-Xval-demographics, echo=F, results="asis",cache=FALSE>>=
adni_69_RIDs = c(295,413,619,685,729,782,938,954,1018,1155,907,981,222,324,448,
                 546,553,572,602,610,814,1341,675,681,731,1130,41,68,70,101,
                 249,293,316,414,698,1206,1222,1339,751,1030,3,5,10,16,22,53,
                 168,183,241,1205,328,1095,991,213,159,337,343,642,753,1109,
                 1217,29,56,1188,1293,149,382,431,1202)
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, subset = VISCODE == 'bl' & RID %in%
               adni_69_RIDs, method = "reverse", test=FALSE, overall=TRUE)
caption = 
  "\\textbf{ADNI-1 cross-validation subset demographics.} 
  CN     - Cognitively Normal. 
  LMCI   - Late-onset Mild Cognitive Impairment. 
  AD     - Alzheimer's Disease.  
  Hisp   - Hispanic. 
  CDR-SB - Clinical Dementia Rating-Sum of Boxes.  
  ADAS   - Alzheimer's Disease Assessment Scale.  
  MMSE   - Mini-Mental State Examination."
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      label="tab:ADNI1-xvalidation-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Sweave/Knitr init

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{article}\usepackage{graphicx, color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\definecolor{fgcolor}{rgb}{0.2, 0.2, 0.2}
\newcommand{\hlnumber}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlfunctioncall}[1]{\textcolor[rgb]{0.501960784313725,0,0.329411764705882}{\textbf{#1}}}%
\newcommand{\hlstring}[1]{\textcolor[rgb]{0.6,0.6,1}{#1}}%
\newcommand{\hlkeyword}[1]{\textcolor[rgb]{0,0,0}{\textbf{#1}}}%
\newcommand{\hlargument}[1]{\textcolor[rgb]{0.690196078431373,0.250980392156863,0.0196078431372549}{#1}}%
\newcommand{\hlcomment}[1]{\textcolor[rgb]{0.180392156862745,0.6,0.341176470588235}{#1}}%
\newcommand{\hlroxygencomment}[1]{\textcolor[rgb]{0.43921568627451,0.47843137254902,0.701960784313725}{#1}}%
\newcommand{\hlformalargs}[1]{\textcolor[rgb]{0.690196078431373,0.250980392156863,0.0196078431372549}{#1}}%
\newcommand{\hleqformalargs}[1]{\textcolor[rgb]{0.690196078431373,0.250980392156863,0.0196078431372549}{#1}}%
\newcommand{\hlassignement}[1]{\textcolor[rgb]{0,0,0}{\textbf{#1}}}%
\newcommand{\hlpackage}[1]{\textcolor[rgb]{0.588235294117647,0.709803921568627,0.145098039215686}{#1}}%
\newcommand{\hlslot}[1]{\textit{#1}}%
\newcommand{\hlsymbol}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlprompt}[1]{\textcolor[rgb]{0.2,0.2,0.2}{#1}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable}
\usepackage{titlesec, siunitx}
\usepackage[round,authoryear]{natbib}
\usepackage[section]{placeins}  % keep floats in their place.
\usepackage{authblk}            % Listing Author affiliations
\usepackage[hypcap]{caption}

@@

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T }
\newcommand{\FEPdataset}{SZ-FEP }

% title and authors
\title{Bootstrapping Multi-atlas Hippocampal Segmentation with \mb}
\author[1]{Jon Pipitone}
\author[1]{Matt T. Park}
\author[1]{Julie Winterburn}
\author[1]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4]{Jens C. Pruessner}
\author[4]{Martin Lepage}
\author[1,5]{Aristotle Voineskos}
\author[1,5,6]{M. Mallar Chakravarty}
\author[ ]{Alzheimer's Disease Neuroimaging Initiative}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health, Hospital for Sick Children, Toronto,
ON,Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Brain Imaging Group, Douglas Hospital Research Centre, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[6]{Institute of Biomedical Engineering, University of Toronto, Toronto,
ON, Canada}

@@

\begin{abstract}
\end{abstract}

@@

The hippocampus is a brain structure situated in the medial temporal
lobe, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}.  The hippocampus is of interest to neuroscientists because it is implicated in
many forms of brain disfunction, including Alzheimer's disease
\citep{Sabuncu2011} and schizophrenia \citep{Narr2004,Karnik-Henry2012}.  In
neuroimaging studies, structural magnetic resonance images (MRI) are often used for the volumetric assessment of the hippocampus.  As such, accurate segmentation of the
hippocampus and its subregions in MRI is a necessary first step to better
understand interindividual variability of neuroanatomy of subjects. Typically, the gold standard for
neuroanatomical segmentation is manual delineation by an expert human rater. However,
the rapid increase in the number of available MRI data and the time and expertise
required for manual segmentation makes this increasingly difficult
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}.
Further, there is significant variation between segmentation protocols regarding specific anatomical boundaries of the 
hippocampus \citep{Geuze2004} and this has
led to efforts to create an unified hippocampal segmentation protocol
\citep{Jack2011}.

Automated segmentation techniques that are reliable, objective, and reproducible
can be considered complementary to manual segmentation.  In the case of classical
model-based segmentation methods \citep{Haller1997,Csernansky1998}, an MRI atlas
that was previously manually labelled by an expert rater is matched to target
images using nonlinear registration methods.  The resulting nonlinear
transformation is applied to the manual labels (i.e. {\em label propagation}) to
move them into the target image space.  While this methodology has been used
successfully in several contexts
\citep{Chakravarty2008,Chakravarty2009,Collins1995,Haller1997}, it is limited
in accuracy by the introduction of errors due inaccuracies in the nonlinear
transformation itself, partial volume effects in label resampling, and
irreconcilable differences between the neuroanatomy represented within the atlas
and target images.

@@

More recently, many groups have been using multiple atlases to improve overall
segmentation accuracy (i.e. multi-atlas segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into a definitive segmentation for the target.  In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such a cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations.  For example \citet{Collins2010}
demonstrated that only 14 atlases, selected based on highest similarity between
medial temporal lobe neuroanatomy as evaluated by normalized mutual information
\citep{Studholme1999} from a library of 80 atlases, were required to achieve
accurate segmentations of the hippocampus.  Several methods have been explored
for label fusion including the STAPLE algorithm \citep{Warfield2004} that
computes a probabilistic segmentation using an expectation maximization
framework from an set of competing segmentations; or others where a subset of
segmentations can be estimated using metrics such as the sum of squared
differences in the regions of interest to be segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library; ranging from atlas libraries
that require between 30 \citep{Heckemann2006} and 80 \citep{Collins2010}
manually segmented atlases.  This strategy has the main drawback of being
inflexible as it does not easily accommodate varying the definition of the
hippocampal anatomy (such as the commonly used heuristic of subdividing the
hippocampus in to head, body, and tail \citep{Poppenk2011,Pruessner2000}).
Furthermore, none of these methods have demonstrated sufficient flexibility to
accommodate atlases that are somehow exceptional such as those derived from
serial histological data \citep{Chakravarty2006,Yelnik2007} or high-resolution MRI
data that enables robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}.  Due to
the recent availability of the latter, there has been increased interest in the
use of probabilistic methods for the identification of the hippocampal subfields
on standard T1-weighted images.  Our group recently demonstrated that through
use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{MallarChakravarty2012}.
The novelty of this manuscript is the extension our multi-atlas methodology to
the hippocampus using more than a single input atlas, while simultaneously
limiting the number of possible inputs used during segmentation, and
demonstrating that accurate identification of the hippocampal subfields is
indeed possible using this methodology.

There are few methods that have attempted to perform multi-atlas segmentation
with a limited number of input atlases. The LEAP algorithm is an elegant
modification to the basic multi-atlas strategy \citep{Wolz2010} in which the
atlas library is grown, beginning with a set of manually labelled atlases, and
successively incorporating unlabelled target images after themselves being
labelling using multi-atlas techniques. The sequence in which target images are
labelled is chosen so that the similarity between the atlas images and the
target images is minimised at each step, effectively allowing for deformations
between very dissimilar images to be broken up into sequences of smaller
deformations.  Although \citet{Wolz2010} begin with an atlas library of 30 MR
images, this method could theoretically work using a much smaller atlas library.
In their validation, LEAP was used to segment the whole hippocampus in the
ADNI-1 baseline dataset, achieving a mean Dice score of 0.85 with manual
segmentations.

To the best of our knowledge there are only two other segmentation methods that
attempt to define the hippocampal subfields using standard T1-weighted data.
The first is included with the FreeSurfer package \citep{VanLeemput2009}.  This
work is limited as it omits the tail of the hippocampus and the segmentation
protocol has yet to be fully validated.  Nonetheless, they demonstrate the
applicability of their work using data from 10 subjects.  In the second method,
\citet{Yushkevich2009} hippocampal subfields were acquired and labelled on
high-resolution MRI data from post-mortem medial temporal lobe samples. Using
nonlinear registration guided by manually derived hippocampus masks and specific
landmarks, the authors demonstrated accurate parcellation of hippocampal subfields in
unlabelled MRI volumes.

Here we address the issue of limiting the number of input atlases by tuning our
algorithm, for segmentation of the entire hippocampus, using a multi-fold
experiment performed on a subset Alzheimer's Disease Neuroimaging Initiative
(ADNI) 1 dataset. Based on the parameters we find in our experiment, we validate
our algorithm using all of the data available in the ADNI Complete 1Yr sample
and compare our segmentations to the other segmentations that are available
through the ADNI informatics portal.  To ensure that we have not over-fit our
parameters to the aging or neurodegenerative brain, we also apply our
segmentations to a dataset of normal controls and individuals suffering from
first episode psychosis. Finally, we perform a leave-one-out validation
experiment to determine if the subfields can be accurately identified using our
multi-atlas framework.

@@

\subsection{The \mb Algorithm}
In this paper, we will use the term {\em label} to mean any segmentation (manual or 
derived) of an MR image.  {\em Label propagation}, or {\em labelling}, is the 
process by which two images are registered and the resulting transformation 
is applied to the labels from one image to bring them into alignment with the 
other image. We will use the term {\em atlas} to mean a manually segmented image,
and the term {\em template} to mean an automatically segmented image (i.e. 
via label propagation). The terms {\em atlas library} and {\em template library} 
describe any set of such images.Additionally, we will use the term {\em target} 
to refer to an unlabelled image that is undergoing segmentation.

The simplest form of multi-atlas segmentation, (which we will call {\em basic 
multi-atlas segmentation}), involves three steps.  First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propagated to the target image space. 
Third, the labels are combined into a single labelling by way of a label fusion method 
\citep{Heckemann2006, Heckemann2011}. This method is described in detail in other
publications \citep{Collins2010,Heckemann2011,Aljabar2009}.

\mb bootstraps the creation of a large template library given a limited input atlas 
library, and then uses the template library in basic multi-atlas segmentation. 
Images for the template library are selected from a set of input target images, 
either arbitrarily or so as to reflect the neuroanatomy or demographics of the 
target set as a whole (for instance, by sampling equally from cases and controls). 
The template library images are then labelled by each of the atlases. Basic 
multi-atlas segmentation is then conducted using the template library
to segment the entire set of target images (including the targets whose images
are used in the construction of the template library).  Since each template
library image has multiple labels (one from each atlas), the final number of 
labels to be fused for each target may be quite large (i.e. \# of atlas $\times$ 
\# of templates).

@@

The following section describes the experiments we conducted to assess the
segmentation quality of the \mb algorithm. The first two experiments assess the
validity of \mb using a cross-validation design. Experiment 1 investigates the
accuracy of \mb whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 tests hippocampal subfield segmentation quality. The last two
experiments assess the validity of the \mb algorithm when applied to different
diseases: first episode schizophrenia (Experiment 3), and Alzheimer's disease
(Experiment 4). Additionally, in experiment 4, we compare \mb segmentations with
those of well-known automated and manual methods and assess segmentation bias.

@@

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a pool
of images and manual hippocampal segmentations from ADNI1 dataset.  This form of
cross-validation allows us to rigorously validate a large number of parameter
settings of \mb (atlas and template library sizes, registration algorithm, and 
label fusion method) and select the best parameters to use in subsequent 
experiments.

\paragraph{ADNI1 dataset}

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Manual segmentations of the left and
right whole hippocampi are available \citep{Hsu2002}. These labels have been
generated using the SNT tool from Medtronic Surgical Navigation Technologies,
Louisville, CO (see Supplementary Materials for detailed discussion of the
manual segmentation process used).

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset download was the "\adnidataset"
standardized dataset available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR
= 2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240
x 240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel resolution of $1.25 \times 1.25 \times 1.2mm^3$.
